---
title: "AE-2"
format: html
editor: visual
---

```{r}

# install(fivethirtyeight)
library(tidyverse)
library(fivethirtyeight)


```

## 🎨 Bob Ross 🎨

Using the `bob_ross` data from `fivethirtyeight` package

1.  How likely was Bob Ross to include a happy little `tree` in one of his paintings?

2.  How has the frequency with which Bob Ross included `clouds` in his paintings changed across the show's seasons? Make a time series to illustrate.

3.  If there is a `mountain` in a Bob Ross painting, how likely is it that mountain is *snowy* (`snowy_mountain`)?

4.  How much more likely was Steve Ross (`steve`) to paint a lake (`lake`) than his dad?Running Code

```{r}

```

## 💺 Flying etiquette 💺

Using the `flying` data from `fivethirtyeight`

1.  In a row of three seats, who should get to use the middle arm rest (`two_arm_rests`)? Make a barplot of the *percent* of respondents who gave each answer.

2.  In general, is it rude to knowingly bring unruly children on a plane? Make a barplot of the *percent* who gave each answer, but **separated by** whether the respondent has a kid or not.

3.  Make a barplot of responses to an etiquette dilemma of your liking. Bonus points if you break it down by a respondent characteristic.

```{r}

```
