---
title: "Lab-1"
output: html_document
---

# Instructions

-   You will do your work in this `Rmarkdown` document but will submit all your answers on Canvas
-   *very important*: As you move through the document, make sure to run all code chunks (grey bits) that already have code in them
-   Write your own code in the empty code chunks
-   Useful shortcuts:
    -   to run all the code in a specific code chunk, press the green right-facing triangle at the top right of the code chunk
    -   to run all *prior* code chunks, press the downward-facing gray triangle at the top right of the code chunk

```{r load-libraries, warning=FALSE, message=FALSE}

library(tidyverse)  # Load ggplot2, dplyr, and all the other tidyverse packages
library(babynames) # loads babynames data, needs to install if not yet installed
library(socviz) # loads socviz data, needs to install if not yet installed
library(juanr) # install if not yet installed
```

# Task 1: Baby Names

The `babynames` package includes data from the Social Security Administration on how often a name was given to a baby between 1880-2015.

Run the code chunk below to create `baby_subset`, a dataset of babies named "Angel".

```{r}
baby_subset <- babynames %>%
  filter(name == "Angel")

# look at the data
head(baby_subset)
```

Using `baby_subset`, make a *time-series* that shows how often the name we picked was used for babies categorized as male and as female over time. (Hint: map the `sex` variable to the color aesthetic; also, the `n` variable shows the count per year).

Make the plot below:

```{r}

```

**Question 1**: Roughly speaking, what happens to the use of the name from the 1990s onwards?

Now, make a similar graph for your name or a name of your choosing. Alter the code chunk below so that `baby_subset` contains babies with your chosen name instead of "Angel". If `baby_subset` looks empty, make sure the name is spelled correctly (e.g., "Juan" instead of "juan"). If you're still struggling, try a different name!

```{r mckinley-subset}
baby_subset <- babynames %>%
  filter(name == "Angel")

# look at the data
head(baby_subset)
```

Now make the same plot from earlier below:

```{r}


```

**Question 2: Briefly describe the trend you observe with your chosen name over time.**

Finally, let's look at the most popular names in 1945. Run the code below:

```{r}
top_names = babynames %>% 
  filter(year == 1945) %>% 
  group_by(sex) %>% 
  top_n(n = 10, wt = prop) %>% 
  select(sex, name, prop) %>% 
  mutate(name = fct_reorder(name, prop))
```

Using `top_names`, make a barplot that puts the baby's name on the y-axis, and the proportion of babies with that name (the variable is called `prop`) on the x-axis. Map the sex variable to the fill aesthetic to create separate bars for male and female babies.

```{r}


```

**Question 3: Roughly speaking, where is the biggest drop off in name popularity among female names?**

# Task 2: Childcare costs

Run the code chunk below to generate `childcare_four`, a dataset of childcare costs for every county in four US states.

```{r}
childcare_four = childcare |> 
  filter(state %in% c("Mississippi", "Virginia", "California", "Hawaii")) |> 
  select(state, county, infant_price)
childcare_four
```

Using `childcare_four`, make a boxplot plot of the distribution of infant childcare prices across states (hint: the state's name should go on the y-axis).

```{r}



```

**Question 4: Generally speaking, which state has the most outliers in terms of childcare costs?**

# Task 3: The Movies

Next, let's look at some data about movies. These data are about a selection of films and info about how much they cost, how much they made in theaters, and what rating they received from reviewers. Run the code below:

```{r clean-movie-data, message=FALSE, warning=FALSE}
# read data
?movies # look at the codebook
movies = 
  movies %>% 
  # exclude films made before the 60s
  filter(year >= 1960)

head(movies)
```

Using `movies`, make a scatterplot that presents a movie's budget (x-axis) against how much it grossed in theaters (y-axis). Color the points by which `decade` the movie was made in. Add titles to the x and y axis and a theme.

Make the plot below:

```{r budget}


```

**Question 5: right click the graph --\> "save image as" --\> and save it somewhere on your computer. You will need to submit the plot!**

Look at that outlier point in top-left part of the plot (should have the color for "1970s"). What movie is this?

Make a new plot that adds labels to each of the points so that you can figure out what that movie is. Hint: look at how we did this with the health and wealth plot from class with `gapminder`.

Make the plot below (hint: if hard to see, press the icon with an arrow and a box at the top right of the image):

```{r}



```

**Question 6: What's the outlier movie?**

**Question 7: What, if anything, does the movie's position on the graph tell us about how successful it was?**

Movies vary so much in length (i.e., their duration in minutes). Around how long are most movies? Make a histogram of `duration` below:

```{r ratings-year}


```

**Question 8: Roughly, about how long are most movies (where is the peak of the distribution)?**

Finally, let's look at who are the most/least *profitable* thriller movie directors. First, we'll make a new dataset that only looks at thriller movies, calculates each movie's profit (`gross` - `budget`) and finds the average profit for each director. To make this more manageable, let's look at directors who made at least 2 movies.

Run the code below to create `directors_profit`, a dataset of thriller movie directors:

```{r summarize-profits-director}
directors_profit = movies %>% 
  # only look at horror movies
  filter(genre1 == "Thriller" | genre2 == "Thriller" | genre3 == "Thriller") %>% 
  # calculate profit
  mutate(profit = gross - budget) %>% 
  # group by director name
  group_by(director) %>% 
  # calculate number of movies each director made, average profit, average rating
  summarise(n_movies = n(), 
            profit = mean(profit)) %>% 
  # include directors who made more than 2 movies
  filter(n_movies > 2)
```

Using `directors_profit` make a barplot that has the `director` on the y-axis and `profit` on the x-axis.

Make the plot below:

```{r plot-director-profits}


```

-   **Question 9: Who is the least profitable director?**

# Task 4: Pokemon

First we'll look at data on pokemon. Let's load it below and look at it.

```{r load-pokemon}
pokemon

?pokemon
```

Using `filter()`, tell me which Pokemon:

-   is taller than a meter
-   weighs more than 200 times its height
-   has a type 1 of "ground"
-   is a generation 3 pokemon

```{r}


```

**Q10: Who's that Pokemon?**

Now, look at Pokemon that:

-   have a higher defense than attack
-   have a capture rate of 3
-   are legendary

```{r}


```

**Q11: How many pokemon fit this description?**

Next, look at Pokemon that are:

-   grass for type 1 and psychic for type 2, or grass for type1 and flying for type2
-   better at speed than defense
-   have an HP level above 50

```{r}


```
