---
title: "Correlations"
---

```{r setup, include=FALSE}
# set knit options
knitr::opts_chunk$set(
  fig.width=9, 
  fig.height=5, 
  fig.retina=3,
  fig.align="center",
  out.width = "100%",
  cache = FALSE,
  echo = FALSE,
  message = FALSE, 
  warning = FALSE
)

# load libraries
library(socviz)
library(juanr)
library(gapminder)
library(patchwork)

# source
source(here::here("slides/R", "funcs.R"))

```

## Quantifying correlation

Correlations are used to understand a relationship between two variables. We use the following equation to quantify correlations:

$$r =
  \frac{ \sum_{i=1}^{n}(x_i-\bar{x})(y_i-\bar{y}) }{\sqrt{\sum_{i=1}^{n}(x_i-\bar{x})^2\sum_{i=1}^{n}(y_i-\bar{y})^2}}$$

-   ranges from -1 to 1,

-   perfectly correlated = 1 or -1,

-   perfectly uncorrelated = 0

All of the movement is here: $\sum_{i=1}^{n}(x_i-\bar{x})(y_i-\bar{y})$

```{r}

child_table = gss_sm |> 
  mutate(age_avg = mean(age, na.rm = TRUE),
         child_avg = mean(childs, na.rm = TRUE),
         diff_age = age - age_avg,
         diff_child = childs - child_avg,
         product = diff_age * diff_child) |> 
  sample_n(5) |> 
  select(age, "\\(\\overline{age}\\)" = age_avg, 
        "\\(\\ age_{i} - \\overline{age}\\)" = diff_age,
         child = childs, "\\(\\overline{child}\\)" = child_avg, 
         "\\(\\ child_{i} - \\overline{child}\\)" = diff_child,
         product) |> 
  slice(c(1, 3, 5))


kbl(child_table, digits = 2, escape = FALSE)
```

$$\sum_{i=1}^{n}(age_i-\overline{age})(child_i-\overline{child})$$

Lots of big, positive **products** $\rightarrow$ strong **positive correlation**

Lots of big, negative **products** $\rightarrow$ strong **negative correlation**

::: callout-important
It is important to understand that correlations **DO NOT** say anything about how much Y changes when X changes. Rather, it only describes the ‘relationship’ between two variables.
:::

Correlations can describe two things: direction and strength of a relationship.

#### **Direction (+/-)**

-   A positive correlation means that when one variable increases another variable increases as well.

-   A negative correlation means that when one variable goes up another variable decreases.

#### **Strength**

-   Correlations range from -1.0 to +1.0.

-   The closer it is to +1 or -1, the more closely the two variables are related. (Strong)

-   If the correlation is close to 0, it means there is no relationship.

Furthermore, to describe correlations we look at the direction (positive or negative) and strength (the number)

-   **Strong:** the relationship is very consistent

-   **Weak:** the relationship exists, but is not very strong or consistent

-   **Positive:** when one variable increases, the other tends to increase

-   **Negative:** when one variable increases, the other tends to decrease

::: callout-warning
**REMINDER: THE NUMERICAL VALUE DOES NOT DESCRIBE HOW MUCH CHANGE JUST THE STRENGTH OF THE RELATIONSHIP.**
:::

### Example 1:

In the first example, we are looking at randomly generated data to see how we can identify correlations visually based on scatter plots. I have provided 5 examples.

```{r}
library(tidyverse)
set.seed(123)

x = rnorm(100, mean = 1, sd = .5)
y = rnorm(100, mean = 1, sd = 0.5)

d1 <- data.frame(x = x,
                 y = y + x*.5)
d2 <- data.frame(x = x,
                 y = y + x*-.5)
d3 <- data.frame(x = x,
                 y = y + x*1)
d4 <- data.frame(x = x,
                 y = y + x*-1)
d5 <- data.frame(x = x,
                 y = y)

# visualize the correlations through a scatter plot

# weak positive correlation
ggplot(d1, aes(x, y)) +
geom_point()  +
 labs(title="Weak Positive Correlation Between X and Y")

# weak negative correlation
ggplot(d2, aes(x, y)) +
geom_point()  +
 labs(title= "Weak Negative Correlation Between X and Y")

# strong positive correlation
ggplot(d3, aes(x, y)) +
geom_point()  +
 labs(title="Strong Positive Correlation Between X and Y")

# strong negative correlation
ggplot(d4, aes(x, y)) +
geom_point()  +
 labs(title="Strong Negative Correlation Between X and Y")

# no correlation
ggplot(d5, aes(x, y)) +
geom_point() +
 labs(title="No Correlation Between X and Y")

```

### Example 2:

Now in example 2, I have randomly generated data on a fake data set to understand how to use the `cor()` function and how to visually plot correlations to illustrate it.

```{r}

age <- sample(18:100, 100, replace = TRUE)
gender <- sample(c(1, 0), 100, replace = TRUE)  # 1 = Female, 0 = Male
kids <- sample(0:5, 100, replace = TRUE)

# Now build the data.frame with dependent vars defined inline
data <- data.frame(
  age = age,
  gender = gender,
  height = rnorm(100, mean = 5.5, sd = 0.75) + gender * -0.25,
  kids = kids,
  income = rnorm(100, mean = 45000, sd = 10000) + age * 1000 + gender * (-2000) + kids * (-3000),
  years_edu = sample(10:23, 100, replace = TRUE) + age * 1 + gender * 2
)

head(data)
```

The `cor()` function allows us to calculate the correlation between two or more variables. My first example looks at only two variables and then visualizes it through a scatter plot.

```{r}

# correlation for only two variables
cor(data$age, data$income)

# visualize the correlations through a scatter plot
ggplot(data, aes(x = age, y = income)) +
geom_point()

```

Since the correlation score is 0.89, we can determine there is a strong positive relationship between age and income, stating that as age increases, so does income, which is seen in our scatter plot as well.

Next, we will examine our full data set. Doing this only works if you have only numerical values; it does not work for non-numerical values.

```{r}

# correlations for all variables in the dataset
cor(data)

```

We can visualize this using the `ggcorrplot()` function.

-   `lab = TRUE` shows the correlation coefficients

-   `type = "upper"` changes the shape of visualization showing

```{r}

# install.packages("ggcorrplot")
library(ggcorrplot) # for ggcorrplot()

# save correlation values as a variable
cor_data <- cor(data)

# plot the correlation values
ggcorrplot(cor_data, lab = TRUE, type = "upper")

```
