---
title: "Extra Materials: Sampling, Bootstraping, and Prediction"
---

```{r, echo = TRUE}


# load libraries
library(socviz)
library(juanr)
library(gapminder)
library(patchwork)
library(broom)
library(paletteer)
library(moderndive)

```



```{r setup, include=FALSE}
# set knit options
knitr::opts_chunk$set(
  fig.width=9, 
  fig.height=5, 
  fig.retina=3,
  fig.align="center",
  out.width = "100%",
  cache = FALSE,
  echo = FALSE,
  message = FALSE, 
  warning = FALSE
)


# source
source(here::here("slides/R", "funcs.R"))
```


## Sampling in R

Now imagine that instead of having data on every American, we only have a **sample** of 10 Americans. Why do we have a sample? Because interviewing every American is prohibitively costly. Same way a poll works: a **sample** to *estimate* American public opinion

We can pick one sample of 10 people from `gss_sm` using the `rep_sample_n()` function from `moderndive`:

```{r, echo = TRUE}
gss_sm %>% rep_sample_n(size = 10, reps = 1)
```

::: callout-note
`size` = size of the sample; `reps` = number of samples
:::

We can then calculate the average number of kids *among that sample of 10 people*

```{r, echo = TRUE}
gss_sm %>% rep_sample_n(size = 10, reps = 1) %>% 
  summarise(avg_kids = mean(childs, na.rm = TRUE))
```

This is our **sample estimate** of the **population parameter**.

## How to simulate

If we take lots of samples of size 20 $\rightarrow$ uncertainty in a sample of 20.

```{r, echo = TRUE, eval = FALSE}
gss_sm %>%
  rep_sample_n(size = 20, reps = 1000)
```

If we take lots of samples of size 100 $\rightarrow$ uncertainty in a sample of 100:

```{r, echo = TRUE, eval = FALSE}
gss_sm %>%
  rep_sample_n(size = 100, reps = 1000)
```

So to see how uncertain we should feel about `gss_sm`, we should take many samples that are **the same size** as `gss_sm`:

```{r, echo = TRUE, eval = FALSE}
gss_sm %>% 
  rep_sample_n(size = 2867, reps = 1000)
```

::: callout-note
`nrow(DATA)` tells you how many observations in data object
:::

If we have a dataset of 2,867 observations and ask R to randomly pick 2,867 observations, we'll just get a bunch of copies of the original dataset:

```{r, echo = TRUE, eval = FALSE}
gss_sm %>% 
  rep_sample_n(size = nrow(gss_sm), reps = 1000)
```

Solution: sample **with replacement** $\rightarrow$ once we draw an observation it goes back into the dataset, and can be sampled again!

```{r, echo = TRUE, eval = FALSE}
gss_sm %>% 
  rep_sample_n(size = nrow(gss_sm), reps = 1000, replace = TRUE)
```

## Sampling with and without replacement

If we were sampling **4** of these delicious fruits:
```{r, echo = TRUE}
fruits = c("Mango", "Pineapple", "Banana", "Blackberry")
fruits
```

It would look like this, with and without replacement:

```{r}
set.seed(1990)
tibble(`Sample, no replace` = sample(fruits, size = 4),
       `Sample, replace` = sample(fruits, size = 4, replace = TRUE)) %>% 
  t() %>% 
  knitr::kable()
  
```

Example: Polling Crime Victimization

```{r, echo = TRUE}
library(juanr)

data <- crime |>
  filter(pais == "Mexico")

boot_data = data |> 
  rep_sample_n(size = nrow(data), reps = 1000, replace = TRUE) |> 
  summarise(avg = mean(vic1ext, na.rm = TRUE),
            standard_error = sd(vic1ext, na.rm = TRUE)) |>
  mutate(low = avg - 1.96 * standard_error, 
         high = avg + 1.96 * standard_error) 

ggplot(boot_data, aes(x = avg)) + 
  geom_histogram(fill = red, color = "white") + 
  theme_nice() + 
  labs(x = "Average number of Individuals who are Victims of a Crime in Mexico") +
  annotate(geom = "rect", 
           xmin = quantile(boot_data$avg, probs = c(.025)), 
           xmax = quantile(boot_data$avg, probs = c(.975)), 
           ymin = 0, ymax = Inf, 
           alpha = .5, fill = yellow)
```


## Bootstraping Regression

We can simulate uncertainty in our estimate by bootstrapping

```{r, echo = TRUE}
boot_lm_kids = gss_sm %>% 
  rep_sample_n(size = nrow(gss_sm), reps = 1000, replace = TRUE) %>% 
  nest(data = -replicate) %>% 
  mutate(model = data %>% map(.x = ., .f = ~lm(childs ~ age, data = .x))) %>% 
  mutate(results = map(model, tidy)) %>% 
  unnest(results) %>% 
  ungroup() %>% 
  filter(term == "age") 

boot_lm_kids %>% 
  select(`Bootstrap` = replicate, `Coefficient` = term, Estimate = estimate) %>%
  slice(1:10) %>% 
  knitr::kable(digits = 4)
```

How much might our estimate of `lm(childs ~ age)` vary? Look at the standard error

```{r, echo = FALSE}
boot_lm_kids %>% 
  summarise(mean = mean(estimate), 
            standard_error = sd(estimate))
```

You might report it is .0345, +/- 2 standard errors

We can also look at the 95% confidence interval -- we are 95% "confident" the effect of age on the number of children a person has is `lm()` will estimate the standard error of regression estimates for us using the **statistical theory** approach, but the results are similar to **boostrapping**

```{r, echo = TRUE}
mod = lm(obama ~ sex, data = gss_sm)
tidy(mod)
```

We can also ask `tidy` for the 95% CI:

```{r, echo = TRUE}
tidy(mod, conf.int = TRUE)
```

`conf.low` and `conf.high` are the lower and upper bound of the 95% CI



## Prediction (by hand)

```{r, echo = TRUE}
weight_model = lm(mpg ~ wt, data = mtcars)
```

The equation for model is:

$\widehat{mpg} = 37.29 - 5.34(wt)$

We only have one explanatory variable to work with (`wt`), so our scenario can only rely on weight

**Scenario**: a car that weights 3.25 tons $\rightarrow$ what fuel efficiency should we expect?

To get an estimate of `mpg`, we simply plug in the value of "weight" we are interested in:

Estimate for weight = [3.25]{.red}

$\widehat{mpg} = 37.29 - 5.34 \times \color{red}{3.25} = 19.935$

## Prediction (in R)

First **define the scenario** we want a prediction for, using `crossing()`

```{r, echo = TRUE}
weight_scenario = crossing(wt = 3.25) 
weight_scenario
```

::: callout-note
the variables in `crossing` have to have the same name as the variables in the model
:::

We can then combine our scenario with our model using `augment()`

```{r, echo = TRUE}
augment(weight_model, newdata = weight_scenario)
```

`augment` takes the values for our scenario and plugs them in to our model equation to get an estimate

::: callout-note
We tell `augment` what our new scenario is using the `newdata =` argument
:::

We can also look at multiple scenarios, maybe a light, medium, and heavy car:

```{r, echo = TRUE}
weight_scenario = crossing(wt = c(1.5, 3, 5)) 
weight_scenario
```

And then use `augment` to get the estimates:

```{r, echo = TRUE}
augment(weight_model, newdata = weight_scenario)
```

## Sequence of scenarios

Or we can look at a *sequence* of scenarios using the `seq()` function

Example: estimates for every weight between 2 and 6 tons, in .1 ton increments

```{r, echo = TRUE}
seq_weights = crossing(wt = seq(from = 2, to = 6, by = .1)) 
seq_weights
```

::: callout-note
`seq` needs a starting point (`from`), an end point (`to`), and an interval (`by`)
:::

And then get predictions for all these scenarios:

```{r, echo = TRUE}
augment(weight_model, newdata = seq_weights)
```


## Muiltiple Regression

```{r, echo = TRUE}

big_model = lm(mpg ~ wt + cyl + hp + am, data = mtcars)

big_scenario = crossing(wt = 3, cyl = 4, hp = 120, am = 1)

augment(big_model, newdata = big_scenario)
```

We can use `seq()` to see what happens when horsepower changes while everything else is left at a fixed value

```{r, echo = TRUE}
varying_hp = crossing(wt = 3, cyl = 4, am = 1, hp = seq(from = 50, to = 340, by = 5))

augment(big_model, newdata = varying_hp)
```

## Visualizing predictions

We could then store our estimate, and use it for plotting

```{r, echo = TRUE, out.width="70%"}
hp_pred = augment(big_model, newdata = varying_hp)
ggplot(hp_pred, aes(x = hp, y = .fitted)) + geom_point()
```
